Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-23 latex/main.tex
Add VirusRecomb cite
Closes #311

\section*{INTRODUCTION}
% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of species~\cite{Felsenstein2004}, with small numbers of widely diverged
samples. In this context, recombination is a nuisance which must be 
carefully accounted for~\cite{Rannala2025recombination},
affecting the inference of species trees~\cite{Posada2002effect}
and biasing downstream
analyses~\cite{Schierup2000consequences,Hedge2014bacterial}.
Modern pathogen datasets are not sparsely sampled from different species,
but consist of tens to hundreds of thousands of samples from 
a single species~\cite{
Shu2017,% GISAID ~270k FLU + others
Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
Apetrei2021hiv,% HIV 1,026,217 sequences
Cryptic2022data,% TB 12k
Dyer2025enterobase% 1 million, various things like E-coli
}.
In contrast to the sophisticated model-based approaches 
used to perform species-level tree 
inference~\cite{Suchard2018bayesian,Bouckaert2019beast},
comprehensive phylogenies of these datasets are typically
built with simpler distance-matrix methods [citation?]
or approximations~\cite{Price2009fasttree}, primarily
due to computational scaling limitations, but also because of
complexity of modelling such datasets.
While several methods have been proposed to incorporate
recombination explicitly into model-based phylogenetic 
inference~\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Muller2022bayesian}, 
scalability 
is a major issue and practical application has been limited.

The data collected during the COVID-19 pandemic as part of global
surveillance fundamentally changed these assumptions. 
At the peak in [month] 202X, there were over X thousand samples 
per day deposited in GISAID\cite{Shu2017}, with a total of 
over 17 million as of Sep 2025. This superabundance of data 
overwhelmed classical phylogenetic methods, both in terms 
of the number of sequences that could be processed, but also
more fundamentally in terms of the underlying assumption of 
retrospective data sparsity. [TODO rephrase]
The UShER method \cite{Turakhia2021} conclusively demonstrated 
the value of simpler models that build a tree iteratively
in real-time, quickly becoming an indispensible element 
of the pandemic response\cite{Hinrichs2024}.

>>>>

\section*{INTRODUCTION}
% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of species~\cite{Felsenstein2004}, with small numbers of widely diverged
samples. In this context, recombination is a nuisance which must be 
carefully accounted for~\cite{Rannala2025recombination},
affecting the inference of species trees~\cite{Posada2002effect}
and biasing downstream
analyses~\cite{Schierup2000consequences,Hedge2014bacterial}.
Modern pathogen datasets are not sparsely sampled from different species,
but consist of tens to hundreds of thousands of samples from 
a single species~\cite{
Shu2017,% GISAID ~270k FLU + others
Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
Apetrei2021hiv,% HIV 1,026,217 sequences
Cryptic2022data,% TB 12k
Dyer2025enterobase% 1 million, various things like E-coli
}.
In contrast to the sophisticated model-based approaches 
used to perform species-level tree 
inference~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
comprehensive phylogenies of these datasets are typically
built with simpler distance-matrix methods [citation?]
or approximations~\cite{Price2009fasttree}, primarily
due to computational scaling limitations, but also because of
the complexity of modelling such datasets.
While several methods have been proposed to incorporate
recombination explicitly into model-based phylogenetic 
inference~\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Muller2022bayesian}, 
scalability 
is a major issue and practical application has been limited.
Software support in explicitly \emph{using} recombination-aware
inference is also extremely limited [TODO add some concrete 
details].

The data collected during the COVID-19 pandemic as part of global
surveillance presents fundamental challenges existing appraoches
in two ways. Firstly, the sheer scale of data collected 
overwhelmed even approximate distance-matrix based methods.
At the peak in [month] 202X, there were over X thousand samples 
per day deposited in GISAID\cite{Shu2017}, with a total of 
over 17 million as of Sep 2025. % TODO check numbers
Regularly reinferring trees de-novo from rapidly 
growing datasets is untenable.
The UShER method \cite{Turakhia2021} conclusively demonstrated 
the value of simpler models that build a tree iteratively
in real-time, quickly becoming an indispensible element 
of the pandemic response\cite{Hinrichs2024}.
The second challenge posed is that recombination is 
an essential element of SARS-CoV-2 
evolution\cite{Jackson2021,VanInsberghe2021,Muller2022bayesian,Focosi2022,
Gutierrez2022,Ignatieva2022,Tamura2023},
with multiple deep recombination events now known to be present in 
genealogy of the global population [cites??].
Although several methods have been proposed to \emph{detect}
recombinant samples~\cite{Zhou2023virusrecom}
